---
title: "Practice Session 9_answers"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---


# Introduction : 

The **Central Limit Theorem** for Sample Means indicate that if the sample size $n$ is large, the distribution of sample means from a population with mean $\mu$ and standard deviation $\sigma$ is approximately normally distributed with mean $\mu$ and standard deviation $\sigma / \sqrt{n}$.


Unfortunately, we find **two difficulties** when trying to use this result in practice. **First**, we generally don't know the standard deviation $\sigma$, so we can just use the standard deviation from the sample, $s$, when estimating the standard error of the sample means.

$$
SE=\frac{s}{\sqrt{n}}
$$
**The second** difficulty  is that when we use the  $SE=\frac{s}{\sqrt{n}}$  to estimate the standard error of the sampling distribution of the means, the distribution is no longer  normal. The result is known as a **t-distribution** with ** degrees of freedom**, df= n-1. 



## 1. Single Mean

Let $\bar{x}$ and $s$ be the sample mean and standard deviation, respectively. $t^*$ is an endpoint chosen from a t-distribution with $df=n-1$. And if the distribution of the population is approximately normal or the sample size is large.


* **CI for a One Mean**

The confidence interval for the population mean, $\mu$, is


$$
\bar{x} \pm t^* \cdot \frac{s}{\sqrt{n}}
$$


* **HT for a One Mean: T- Test**

To test $H_{\mathrm{O}}: \mu=\mu_{\mathrm{O}}$ vs $H_a: \mu \neq \mu_{\mathrm{O}}$ (or a one-tail alternative) use the t -test statistic as follows: 

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\bar{x}-\mu_0}{s / \sqrt{n}}
$$

where  the p -value of the test is computed using the appropriate tail(s) of a t -distribution with $n-1$ degrees of freedom.





# 2. Difference in Means

Consider two population approximately normally distributed or the two sample sizes are large. Where $\bar{x}_1$ and $\bar{x}_2$ are the means, $s_1$ and $s_2$ are the standard deviations, and the samples of sizes $n_1$ and $n_2$ respectively. 


 * **CI for a Difference in Means**
 
The confidence interval for the difference in two population means, $\mu_1-\mu_2$, is

$$
\left(\bar{x}_1-\bar{x}_2\right) \pm t^* \cdot \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}
$$

where **$t^*$** is an endpoint chosen from a t-distribution with df = min ( $n_1-1$, $n_2-1$ ) .






* **HT Two-Sample t-Test for a Difference in Means**

To test $H_{\mathrm{O}}: \mu_1=\mu_2$ vs $H_a: \mu_1 \neq \mu_2$ (or one-tail test), we use two-sample t -statistic: 

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\left(\bar{x}_1-\bar{x}_2\right)-0}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}
$$
we use a t -distribution to find the p -value  with df = min ( $n_1-1$, $n_2-1$ ). 



# 3. Inference for a Difference in Means with **Paired Data**

* **CI for Difference in Means with Paired Data** 

To estimate the difference in means based on paired data, we first subtract to compute the difference for each data pair and compute the mean $\bar{x}_d$, the standard deviation $s_d$, and the sample size $n_d$ for the sample differences.Provided the differences are reasonably normally distributed (or the sample size is large).

The  confidence interval for the difference in means is given by:

$$
\text { Statistic } \pm t^* \cdot S E=\bar{x}_d \pm t^* \frac{s_d}{\sqrt{n_d}}
$$

where $t^*$ is a percentile from a t -distribution with $n_d-1$ degrees of freedom.


* **HT for Difference in Means with Paired Data** 

To test $H_{\mathrm{o}}: \mu_d=\mathrm{o}$ vs $H_a: \mu_d \neq \mathrm{o}$ (or one-tail test) we use the t-test statistic:

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\bar{x}_d-0}{s_d / \sqrt{n_d}}
$$


If the differences are reasonably normally distributed (or the sample size is large), we use a t-distribution with $n_d-\mathbf{1}$ degrees of freedom to compute a p -value for this test statistic.









## Exercise2: Single Sample T-Test



According to a local statistics instructor, the average score on his first exam of the semester (`Exam1`) is historically around 80%. The data is available in the `StatsGrades` from the `Lock5Data` library. Run a one sample t-test to see if this year's students performed significantly better than the historic average.

```{r}
library(SDS1000)
library(Lock5Data)
data(StatGrades)
```


f.  Use the `dt()` function and `plot()` to visualize the t-distribution. The degrees of freedom will be the same as for the test. Add a vertical red line to the plot at the observed sample mean.




a. First visualize the data using a boxplot and a histogram. Do you see any outliers?

```{r}
par(mfrow = c(1, 2))

  exam1 <- StatGrades$Exam1

boxplot(exam1,
        main= " boxplt of Exam 1",
        ylab = "Exam 1 Scores")

hist(exam1, xlab = "Exam 1 Scores")
```

There appears to be one outlier. We might want to investigate this score further.

b. State the null and alternative hypotheses using symbols.

* $H_0: \mu = 80$
* $H_A: \mu \neq 80$



c. Calculate the sample statistic.

```{r}

n1 <- length(exam1) # get sample size (n = 50)

# compute sample mean

mean_exam1 <-  mean(exam1) 

```



d. Calculate the t-test statistic

```{r}

# first compute standard error

sd_exam1 <-  sd(StatGrades$Exam1)
se_exam1 <-  sd_exam1 / sqrt(n1) 

# t-test statistic

t_stat_exam1 <- (mean_exam1 - 80) / se_exam1
t_stat_exam1


```


e. Calculate the p-value of the test, and state your conclusion.


```{r}
# df = 50 - 1 = 49
pt(T_stat_exam1, df = n1 - 1, lower.tail = F)
```

**Since our p-value is greater than 0.05, we will fail to reject the null hypothesis. We therefore do not have evidence that this year's exam scores are significantly higher than 80.**

e)

```{r}
# Create sequence of t-values
t_values = seq(-4, 4, length.out = 1000)
dt_values = dt(t_values, df = 49) # convert to density values

# plot and make line plot with type = "l"
plot(t_values, dt_values, type = "l") 
abline(v = T_stat_exam1, col = "red") # add vertical red line
```

**It seems that the sample mean wasn't very far from the null value of 80, which supports the lack of significance in our test.**

# Question 3: Two Sample T Test

a)

```{r}
boxplot(StatGrades$Exam1, StatGrades$Exam2,
        names = c("Exam 1", "Exam 2"),
        main = "Exam Scores",
        ylab = "Score")
```

**The median scores appear to be similar for both exams. Exam 1 scores appear to have more variability than exam 2 scores. There are two outliers in the exam 2 scores.**

b)

* $H_0: \mu_1 = \mu_2$
* $H_A: \mu_1 \neq \mu_2$

c)

```{r}
n2 = length(StatGrades$Exam2) # get sample size (n = 50)

mean_exam2 = mean(StatGrades$Exam2) # compute sample mean
sd_exam2 = sd(StatGrades$Exam2)

# Compute standard error for two means
se_two_means = sqrt(sd_exam1^2/n1 + sd_exam2^2/n2)
T_stat_two_means = (mean_exam2 - mean_exam1) / se_two_means
T_stat_two_means
```

d) 

**Since we are doing a two-sided test, we need to multiply the p-value by 2.**

```{r}
2 * pt(T_stat_two_means, df = 50 - 1, lower.tail = F)
```

**Since our p-value is greater than 0.05, we will fail to reject the null hypothesis. We therefore do not have evidence that this year's students performed differently on the two exams.**

e)

* $H_0: \mu_1 = \mu_{final}$
* $H_A: \mu_1 < \mu_{final}$

**Note that we are doing a one-sided test, with the alternative being improvement on the final compared to the first exam.**

```{r}
n_final = length(StatGrades$Final) # get sample size (n = 50)

mean_final = mean(StatGrades$Final) # compute sample mean
sd_final = sd(StatGrades$Final)

# Compute standard error for two means
se_one_final = sqrt(sd_exam1^2/n1 + sd_final^2/n_final)
T_stat_one_final = (mean_final - mean_exam1) / se_one_final
T_stat_one_final
```

```{r}
pt(T_stat_one_final, df = 50 - 1, lower.tail = F)
```

**Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that student scores improved on the final compared to the first exam.**

# Question 4: Paired T Test

a)

**Let $D$ be the true difference between the exam 1 and exam 2 scores. We will perform the following two-sided test:**

* $H_0: D = 0$
* $H_A: D \neq 0$

b)

```{r}
diff_exams = StatGrades$Exam2 - StatGrades$Exam1
```

c)

```{r}
n_diff = length(diff_exams)

mean_diff = mean(diff_exams)
se_diff = sd(diff_exams) / sqrt(n_diff)
```

d)

```{r}
T_stat_diff = mean_diff / se_diff
```

e)

```{r}
2 * pt(T_stat_diff, df = n_diff - 1, lower.tail = F)
```

**Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that this year's students performed differently on the two exams.**


# *Part 2*  Permutations tests for one mean and diffrence between two means:

## Test for the diffrence between two means with permuations:


The same local statistics instructor believes that students perform the same on the first exam (`Exam1`) and the second exam (`Exam2`). Again, the data is available in the `StatsGrades` data set from the `Lock5Data` library. 

1.) Create a null distribution and create its histogram.

2.) Calculate the p-value


```{r}
n=50 
library(SDS1000)
combined_data <- c(StatGrades$Exam1, StatGrades$Exam2)
null_dist <- do_it (10000) * {
shuff_data <- shuffle(combined_data)
shuff_grp1 <- shuff_data[1:50]
shuff_grp2 <- shuff_data[51:100]
mean(shuff_grp2) - mean(shuff_grp1)
}
```



```{r}
hist(null_dist)

obs_stat <- mean_exam1 -mean_exam2

abline(v = obs_stat, col = "blue")
```


4.)  Calculate p-value 

```{r}

p_value <- pnull(obs_stat, null_dist, lower.tail = T) 
p_value

2*p_value 
# 0.3072
```


## Test for one mean:

**(Let go back to he the question of testing one mean)**.

According to a local statistics instructor, the average score on his first exam of the semester (`Exam1`) is historically around 80%. The data is available in the `StatsGrades` data set from the `Lock5Data` library. let Create a bootstrap distribution with a mean of the hypothesised value

1.) Calculate the mean of the data.

```{r}
mean_exam1 <-  mean(StatGrades$Exam1) # compute sample mean
```

2.) Subtract this mean from the data.

```{r}
subtract_mean <- StatGrades$Exam1- mean_exam1
```

3.) Add the hypothesized value to the data.

```{r}

modified_data <- subtract_mean + 80

mean(modified_data )
```

4.)Now create a bootstrap distribution with the last modified data you have created in the previous step.


```{r}
n=50
my_sample <- modified_data

boot_sample <- sample(my_sample, n, replace = TRUE) #sample with replacement
null_dist  <- do_it(10000) * {
curr_boot <- sample(my_sample, n, replace = TRUE)
mean(curr_boot)
} 


hist( null_dist )

obs_stat <- mean_exam1

abline(v = obs_stat, col = "blue")
```

5.) Calculate the p-value
```{r}
p_value <- pnull(obs_stat, null_dist, lower.tail = F ) 
p_value


```
SDS1000/ClassMaterial/practice_sessions/practice_session_09/practice_09_answers.qmd at main · emeyers/SDS1000

**It seems that the sample mean wasn't very far from the null value of 80, which supports the lack of significance in our test.**

# Two Sample T Test

a)  

```{r}
boxplot(StatGrades$Exam1, StatGrades$Exam2,
        names = c("Exam 1", "Exam 2"),
        main = "Exam Scores",
        ylab = "Score")
```

**The median scores appear to be similar for both exams. Exam 1 scores appear to have more variability than exam 2 scores. There are two outliers in the exam 2 scores.**

b)  

-   $H_0: \mu_1 = \mu_2$
-   $H_A: \mu_1 \neq \mu_2$

```{r}
n2 = length(StatGrades$Exam2) # get sample size (n = 50)

mean_exam2 = mean(StatGrades$Exam2) # compute sample mean
sd_exam2 = sd(StatGrades$Exam2)

# Compute standard error for two means
se_two_means = sqrt(sd_exam1^2/n1 + sd_exam2^2/n2)
T_stat_two_means = (mean_exam2 - mean_exam1) / se_two_means
T_stat_two_means
```

**Since we are doing a two-sided test, we need to multiply the p-value by 2.**

```{r}
2 * pt(T_stat_two_means, df = 50 - 1, lower.tail = F)
```

**Since our p-value is greater than 0.05, we will fail to reject the null hypothesis. We therefore do not have evidence that this year's students performed differently on the two exams.**

c)  

-   $H_0: \mu_1 = \mu_{final}$
-   $H_A: \mu_1 < \mu_{final}$

**Note that we are doing a one-sided test, with the alternative being improvement on the final compared to the first exam.**

```{r}
n_final = length(StatGrades$Final) # get sample size (n = 50)

mean_final = mean(StatGrades$Final) # compute sample mean
sd_final = sd(StatGrades$Final)

# Compute standard error for two means
se_one_final = sqrt(sd_exam1^2/n1 + sd_final^2/n_final)
T_stat_one_final = (mean_final - mean_exam1) / se_one_final
T_stat_one_final
```

```{r}
pt(T_stat_one_final, df = 50 - 1, lower.tail = F)
```

**Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that student scores improved on the final compared to the first exam.**

# Paired T Test

**Let** $D$ be the true difference between the exam 1 and exam 2 scores. We will perform the following two-sided test:

-   $H_0: D = 0$
-   $H_A: D \neq 0$

```{r}
diff_exams = StatGrades$Exam2 - StatGrades$Exam1
n_diff = length(diff_exams)

mean_diff = mean(diff_exams)
se_diff = sd(diff_exams) / sqrt(n_diff)

T_stat_diff = mean_diff / se_diff
```

```{r}
2 * pt(T_stat_diff, df = n_diff - 1, lower.tail = F)
```

**Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that this year's students performed differently on the two exams.**


# *Part 2*  Permutations tests for one mean and diffrence between two means:

## Test for the diffrence between two means with permuations:


The same local statistics instructor believes that students perform the same on the first exam (`Exam1`) and the second exam (`Exam2`). Again, the data is available in the `StatsGrades` data set from the `Lock5Data` library. 

1.) Create a null distribution and create its histogram.

2.) Calculate the p-value


```{r}
n=50 
library(SDS1000)
combined_data <- c(StatGrades$Exam1, StatGrades$Exam2)
null_dist <- do_it (10000) * {
shuff_data <- shuffle(combined_data)
shuff_grp1 <- shuff_data[1:50]
shuff_grp2 <- shuff_data[51:100]
mean(shuff_grp2) - mean(shuff_grp1)
}
```



```{r}
hist(null_dist)

obs_stat <- mean_exam1 -mean_exam2

abline(v = obs_stat, col = "blue")
```

4.)  Calculate p-value 

```{r}

p_value <- pnull(obs_stat, null_dist, lower.tail = T) 
p_value

2*p_value 
# 0.3072
```


## Test for one mean:

**(Let go back to he the question of testing one mean)**.

According to a local statistics instructor, the average score on his first exam of the semester (`Exam1`) is historically around 80%. The data is available in the `StatsGrades` data set from the `Lock5Data` library. let Create a bootstrap distribution with a mean of the hypothesised value

1.) Calculate the mean of the data.

```{r}
mean_exam1 <-  mean(StatGrades$Exam1) # compute sample mean
```

2.) Subtract this mean from the data.

```{r}
subtract_mean <- StatGrades$Exam1- mean_exam1
```

3.) Add the hypothesized value to the data.

```{r}

modified_data <- subtract_mean + 80

mean(modified_data )
```

4.)Now create a bootstrap distribution with the last modified data you have created in the previous step.


```{r}
n=50
my_sample <- modified_data

boot_sample <- sample(my_sample, n, replace = TRUE) #sample with replacement
null_dist  <- do_it(10000) * {
curr_boot <- sample(my_sample, n, replace = TRUE)
mean(curr_boot)
} 


hist( null_dist )

obs_stat <- mean_exam1

abline(v = obs_stat, col = "blue")
```

5.) Calculate the p-value
```{r}
p_value <- pnull(obs_stat, null_dist, lower.tail = F ) 
p_value


```
